---
title: "AKu_MiniAnalysis_1"
author: "Andrea_Ku"
date: "07 October 2021"
output: github_document

---


# **Milestone 1- Mini Data Analysis Project** 

The purpose of this mini data analysis is to learn how to work through a data analysis while creating a clear markdown file. 

Install and load the packages

```{r}
#install.packages("devtools")
#devtools::install_github("UBC-MDS/datateachr")

library(datateachr)
library(tidyverse)
```

# **Task 1- Choose your favorite dataset**

**1.1 Choose 4 of the datateachr packages that appeal to you**

 1. building_permits
 
 2. flow_sample
 
 3. steam_games
 
 4. vancouver_trees
 
**1.2 Explore the dataset**

Explore **building_permits** dataset

What dataframe?
```{r}
class(building_permits)
```
What dimensions in the data frame?
```{r}
dim(building_permits)
```
How many rows is the data? To guage data complexity
```{r}
nrow(building_permits)
```
How many columns is the data? To guage data complexity
```{r}
ncol(building_permits)
```
Print the data
```{r}
glimpse(building_permits)
```

Explore **flow_sample** dataset

What dataframe?
```{r}
class(flow_sample)
```
What dimensions in the data frame?
```{r}
dim(flow_sample)
```
How many rows is the data? To guage data complexity
```{r}
nrow(flow_sample)
```
How many columns is the data? To guage data complexity
```{r}
ncol(flow_sample)
```
Print the data
```{r}
glimpse(flow_sample)
```

Explore **steam_games** dataset

What dataframe?
```{r}
class(steam_games)
```
What dimensions in the data frame?
```{r}
dim(steam_games)
```
How many rows is the data? To guage data complexity
```{r}
nrow(steam_games)
```
How many columns is the data? To guage data complexity
```{r}
ncol(steam_games)
```
Print the data
```{r}
glimpse(steam_games)
```

Explore **vancouver_trees** dataset

What dataframe?
```{r}
class(vancouver_trees)
```
What dimensions in the data frame?
```{r}
dim(vancouver_trees)
```
How many rows is the data? To guage data complexity
```{r}
nrow(vancouver_trees)
```
How many columns is the data? To guage data complexity
```{r}
ncol(vancouver_trees)
```
Print the data
```{r}
glimpse(vancouver_trees)
```

**1.3 Narrow down to 2**

From the selected dataframes, I am most interested in the vancouver_trees dataset and the flow_sample datasets. After reviewing the data that each set contains, I found the flow_sample dataset to be interesting, relevant, and familiar for me, as my interest is in environmental studies. My previous projects looks at guage station data, so the information provided in the flow_sample was more intuitive for me to understand. Similarly, I also chose the vancouver_trees dataset because I think tree identification and measurements are similar to the kind of data I will be working with in my field. Data about trees also interest me because I will likely continue to work with trees moving forward into my career, so the vancouver_trees dataset is both familiar and relevant for me. 

**1.4 Potential research question with each dataset**

**Potential question for the flow_sample dataset:**

What is the relationship between month and flow? How do the flow patterns vary each month?

**Potential question for the vancouver_trees dataset:**

How does the date that the tree was planted relate to the neighborhood? Were the measured trees planted in the same neighborhood at the same time, or were the planted in neighborhoods sporadically? What is the distribution of diameters measured?

**Final Decision:** I've decided to use the flow_sample dataset for my Mini Data Analysis. The flow_sample dataset is more straightforward for me to understand, and is just as relevant for my future work in the environmental sector.

**Task 2: Exploring your dataset**

**2.1 Dive deeper into your data**

1. Plot the distribution of a numeric variable:
Plot the distribution of when the data was taken to see if there were years or months that more data was taken than others. Use the jitter as geometry so we can better distinguish the distribution. 

```{r}
sample_dist <- ggplot(flow_sample) +
    geom_jitter(aes(x = year, y = month), alpha = 0.1)
print(sample_dist)
```

2. Explore the relationship between 2 variables in a plot:
Plot the minimum and maximum flows over months in seperate plots to see the range of flow within each month for each extreme. Facet the plot to seperate the minimum and maximum, so we can more effectively see the range of values

```{r}
Extreme_Flow_Month <- ggplot(flow_sample,
                    aes(month, flow)) +
                    facet_wrap(~ extreme_type, scales = "free_y")+
 geom_point(aes(color = extreme_type),
            size = 3,
            alpha = 0.8)
            guides(color = FALSE)
print(Extreme_Flow_Month)
```

3. Filter observations according to time, where we can look at recent data starting in 1960, and before 1960. Then plot the observations to see if flow has changed over the century.
Filter for recent years, from 1960 till present

```{r}
MaxSince_1960 <- flow_sample %>%
  filter(year > 1960, extreme_type == "maximum")
print(MaxSince_1960)
```

Plot the flow for years 1960-present

```{r}
MaxSince_1960_plot <- ggplot(MaxSince_1960) +
    geom_point(aes(x = year, y = flow), alpha = 0.1)
print(MaxSince_1960_plot)
```

Filter for historical years, since before 1960 

```{r}
MaxBefore_1960 <- flow_sample %>%
  filter(year < 1960, extreme_type == "maximum")
print(MaxBefore_1960)
```

Plot the flow for historical years, since before 1960

```{r}
MaxBefore_1960_plot <- ggplot(MaxBefore_1960) +
    geom_point(aes(x = year, y = flow), alpha = 0.1)
print(MaxBefore_1960_plot)
```


4. Plot the flow in boxplots to see the distribution within minimum and maximum flows
Filter by Maximum Flow

```{r}
Max_only <- flow_sample %>%
  filter(extreme_type == "maximum")
print(Max_only)
```

Plot the distribution of Max flow by Box plot

```{r}
Max_Box <- ggplot(Max_only, aes(extreme_type, flow)) + 
    geom_boxplot(width = 0.2)
print(Max_Box)
```

Filter by minimum flow

```{r}
Min_only <- flow_sample %>%
  filter(extreme_type == "minimum")
print(Min_only)
```

Plot the distribution of Min flow by box plot

```{r}
Min_Box <- ggplot(Min_only, aes(extreme_type, flow)) + 
    geom_boxplot(width = 0.2)
print(Min_Box)
```

**2.2 Explain the exercise in relation to flow_samples**

**1. For plotting the numeric values:**

I will plot the distribution of the data over each month and year the data was taken to see when the most or least data was collected. This will help me determine the range of time that that data was collected, which would help me better understand the data collection and how it may influence my analysis. This plot answers questions like, "Were there equal number of samples collected each month, and each year?" and "Were there more samples collected in the summer or the winter of each year?" The answers to these will help me better analyze the data in case I need to transform or reduce the data into averages, and recognize when the data might be skewed due to fewer samples.

**2. For exploring the relationship between 2 variables in a plot:**

I thought it would be interesting to distinguish the range of flow for both the minimum and maximum extreme over the course of the year.This will help me determine the range of time that that data was collected in the dataset, and get a better sense of the environment I am working in and when the flow seasons are. According to the plot, it seem like month 5-7 have peak flow, which is May, June, and July. This is to show 1) which month are there minimums and maximum, and 2) within each extreme type, what is the range in the extreme? What is the range of the minimum flow? And what is the range of the maximum flow? Are there months where the maximum flow have a greater range? When are the months where the minimum flows have the least range? These questions can help be explained by exploring the relationship in these variables.

**3. For filtering the observations according to time**

The amount of flow may have changed over time, as the climate, land use, or development has changed. Filtering helps to divide the data into different categories that can be used to distinguish if the flow has experienced any changed over the century. Since this data starts in 1909, I thought that dividing the data in half by time would be appropriate to see general trends in flow change. Therefore, historical data would be considered before 1960, and recent data is considered since 1960. The data is filtered by these parameters, then plotted so I can see the range in values over the years. This filtering and plotting method answers questions related to change over time. How does the range of flow since before 1960 compare to the range of flow from 1960 to present day? Has flow generally decreased over time?

**4. For creating boxplots of minimum and maximum flows**

Within each extreme type, there is variability in the data. The box plot is helpful for plotting the variability in the data to show where most of the minimum values lie within the dataset, and likewise for the maximum values. This boxplot helps identify any flow values that are not within the normal range for minimum and maximum values. So if there were any flow values that were abnormally high, I could identify them more easily and either investigate or exclude them from my analysis.

# **Task 3: Write Research Questions**

1. How does the range of flow since before 1960 compare to the range of flow from 1960 to present day? Has flow generally decreased over time?

2. Within each extreme type, how does the pattern of flow change over the course of the month? How does the flow in the first few days of the month compare to the last days of the month?

3. Which months are there minimums and maximum? Does the flow correspond to seasons?

4. Within each extreme type, what is the range in the extreme? Are there any values that are outside the normal distribution of the data? Do these values lie within certain time frames?



# 