---
title: "Mini Data Analysis Milestone 2"
output: github_document
---

# Welcome back to Mini Data Analysis Part 2!

Begin by loading data and the tidyverse package below: 

```{r, message = FALSE}
library(datateachr) # <- might contain the data you picked!
library(tidyverse)
library(ggplot2)
library(ggridges)
```

# Task 1: Process and summarize your data

### 1.1 Research Questions:

Note: The research questions were slightly modified from Milestone 1 to suit the scope of Milestone 2. 

1. How does the range of maximum flow since before 1960 compare to the range of flow from 1960 to present day?

2. Do minimum and maximum extreme flow types correspond to specific seasons?

3. Within the maximum extreme type, how does the pattern of flow change over the seasons?

4. Are there changes in maximum flow patterns by decade?


### 1.2 Summarize and Graph

*1. How does the range of maximum flow since before 1960 compare to the range of flow from 1960 to present day?*

#### Summarizing

##### Create a new column to seperate data for years prior to 1960 and since 1960 for maximum extremes. 
```{r}
flow_sample_PeriodMax <- flow_sample %>% 
  filter(extreme_type == 'maximum') %>%  
  mutate(Period = case_when(year > 1960 ~ "Since_1960",
                             year < 1960 ~ "Before_1960",
                             TRUE ~ "UNK"))
print(flow_sample_PeriodMax)

```

##### Compute the range, mean, median, and standard deviation across the groups of the categorical variable, Period (since_ or before_1960), from your data. 

###### This task helps answer the research question by providing the range of values for the two different time periods, and allows us to easily quantify the change in maximum flow over time.

```{r}
FlowPeriod_summaryMax <- flow_sample_PeriodMax %>% 
              group_by(Period) %>%
  summarise(range = range(flow),
            mean = mean(flow),
            min = min(flow),
            max = max(flow),
            median = median(flow),
            stdDev = sd(flow))
print(FlowPeriod_summaryMax)

```

#### Graphing

##### Create a graph out of summarized variables that has at least two geom layers

###### For this graph, it makes more sense to create the graph from all the variables to show the summarized variables in a boxplot. Previously, the maximum flow values were quantified using values in a table, which is valuable for comparing actual numbers. Now, the summary statistics are displayed in a graph, which is more helpful for seeing general trends in the two categories. As you can see in the boxplot, the flow summary statistics from before 1960 seems generally higher than the flow since 1960, suggesting that the flow has decreased over time.

```{r}
Max_Box <- ggplot(flow_sample_PeriodMax, aes(Period, flow)) + 
    geom_boxplot(width = 0.2)+
    geom_jitter(alpha = 0.5, width = 0.2)
print(Max_Box)

```

*2. Do minimum and maximum extreme flow types correspond to specific seasons?*

#### Summarizing

##### Compute the number of observations for each extreme type by season. 

###### Seasons align with the months for the northern hemisphere, where winter months are December - January, and Summer months are June - August.  

```{r}
# Create new column for each season
flow_sample_Seasons <- flow_sample %>% 
   mutate(season = case_when(month < 3 ~ "winter",
                                 month < 6 ~ "spring",
                                 month < 9 ~ "summer",
                                 month < 12 ~ "autumn",
                                 TRUE ~ "winter"))
print(flow_sample_Seasons)
```

##### The observations of each extreme type is tallied by season, which identifies how many times each extreme observation occured in each season.

###### This information addresses the research question by highlighting whether minimum or maximum extremes are noted in specific seasons. For example, maximum extremes were only observed in spring and summer, suggesting that spring and summer have heavier water flow than winter and autumn. With this information, we know that maximum extreme flow values do coorespond to specific seasons- spring and summer. Alternatively, the minimum extreme type is observed in autumn, spring, and winter. Therefore, we know that minimum flows are not observed in the summer. 

```{r}
# Count the observations of each extreme type by season
nExtremeType <-  flow_sample_Seasons %>%
        group_by(extreme_type, season) %>%
        tally()
print(nExtremeType)
```

#### Graphing

##### Create a graph out of summarized variables that has at least two geom layers. 

###### The chart uses the summarized counts for minimum and maximum extremes for each season, which quantifies how many times an extreme observation is observed in the seasons. The research question is addressed in the graph, as the maximum observations are only seen in summer and spring months, where the large majority of maximum values are from the summer months. Alternatively, the majority of minimum values are observed in winter months, with spring having the second most minimum observations. Therefore, from the graph, you can clearly see that maximum and minimum extremes do share a relationship with the seasons.

```{r}
# Graph the count (n) for each extreme type, per season
nChart <- ggplot(data = nExtremeType, aes(x = extreme_type)) + 
    geom_col(aes(x= extreme_type, y = n, fill = season), position = "dodge") +
    geom_text(aes(y = n, label = n, group = season),
    colour = "black", size = 4,
    vjust = 1.5, position= position_dodge(0.9))
print(nChart)
```

 
*3. Within the maximum extreme type, how does the pattern of flow change over the seasons?*

#### Summarizing

##### Based on two categorical variables- Maximum extreme type and seasons, calculate two range, mean, median, and standard deviation.

###### The research question is addressed when running the summary statistics by season and by extreme type, as the output table describes the quantitative values of maximum flow in the two seasons that maximum flow occurs- in the spring and summer.

###### From the calculated statistics, you can see how the median and mean of the maximum flow is greater in the summer, suggesting overall greater flow during summer months. However, you will also notice that the range and the standard deviation is also greater in the summer months, which indicates greater variation of flow, meaning that there likely are greater variations in weather patterns at this station during the summer, and more consistent flow patterns during the spring. 

```{r}
# Calculate Statistics
FlowSeasons_summaryMax <- flow_sample_Seasons %>% 
  filter(extreme_type == 'maximum') %>% 
              group_by(season) %>%
  summarise(range = range(flow),
            mean = mean(flow),
            min = min(flow),
            max = max(flow),
            median = median(flow),
            stdDev = sd(flow))
print(FlowSeasons_summaryMax)
```

#### Graphing 

##### Make a graph where it makes sense to customize the alpha transparency. 

###### The density graph is useful for describing the patterns in maximum flow over the season, as you are able to observe the distribution of flow in each season. By observing the graph, the spring months seem to have a higher density of flow between a narrower range of flow measurements, while the summer months seem to have a wider range of flow values. 

###### The alpha transparency in this graph is helpful to when comparing the density of flow patterns over the summer and spring seasons, where there are maximum flow values. Without the alpha transparency, it would be difficult to see the overlap and curavture of the graph for each season. 

```{r}
#Filter by Maximum only, since we only want to know how maximum flow changes over the seasons
flow_sample_SeasonsMax <- flow_sample_Seasons %>% 
  filter(extreme_type == 'maximum')
print(flow_sample_SeasonsMax)

# Plot the density graph for maximum flow over the season
Season_Density <- ggplot(flow_sample_SeasonsMax, aes(x = flow)) +
    geom_density(aes(fill = season), alpha =0.3)
print(Season_Density)
```


*4. Are there changes in maximum flow patterns by decade? *

#### Summarizing

##### Create a decade variables to categorize the flow values by decade for at least the past 30 years. 

###### By sorting the years into individual decades, we can evaluate how the overall flow patterns have changed by comparing the values in each decade. 

```{r}
# Create a new column with at least the past 30 years categorized into decades
flow_sample_decade <- flow_sample_Seasons %>% 
    filter(extreme_type == 'maximum') %>%  # Only want maximum flow values
  mutate(decade = case_when(year > 2009 ~ "2010_2019",
                            year > 1999 ~ "2000_2009",
                            year > 1989 ~ "1990_1999",
                            year > 1979 ~ "1980_1989",
                            year > 1969 ~ "1970_1979",
                            year > 1959 ~ "1960_1969",
                            year > 1949 ~ "1950_1959",
                            year > 1939 ~ "1940_1949",
                            year > 1929 ~ "1930_1939",
                            year > 1919 ~ "1920_1929",
                            year > 1909 ~ "1910_1919",
                            year > 1899 ~ "1900_1909",
                            ))
print(flow_sample_decade)
```

#### Graphing

##### Make a graph where it makes sense to customize the alpha transparency. 

###### To see if there are changes in the maximum flow pattern by decade, the density ridges graph shows the range of flow values over time. I added quantile metrics so the data is easier to compare and contrast the maximum flow patterns between the decades. 

###### The alpha transparency helps visualize the graph and the quantiles, as it focuses the graph on the dark lines, and helps display any quantile lines that overlap with the ridge graph shadowing.

```{r}
# Visualize the flow pattern between each recent decade
chart_decade_flowpatterns <- flow_sample_decade %>% 
   ggplot(aes(flow, decade)) +
 stat_density_ridges(quantile_lines = TRUE, quantiles = c(0.025, 0.975), alpha = 0.3) #Add quantiles
print(chart_decade_flowpatterns)
```

### 1.3 Results and Review of research questions

Based on the operations, I have answered all of my research questions, but each question can be expanded and researched more thoroughly. For example, in question 4, "Are there changes in maximum flow patterns by decade?" the answer could also be investigated by using summary statistics by calculating the mean, range, and standard deviation for each decade. Furthermore, for some of the questions, I only asked about the maximum extreme type. To provide a more comprehensive review of the data, I could also explore the minimum extreme type. Rather than my research questions being refined, I think expanding my investigation would yield interesting results, particularly in regard to exploring the data for minimum flows. 

The research questions that are yielding the most interesting results include question 1 and 2: 1) How does the range of maximum flow since before 1960 compare to the range of flow from 1960 to present day? and 2) Do minimum and maximum extreme flow types correspond to specific seasons? For the range of maximum flow from before 1960 compared to 1960 to present day, I found that the flow summary statistics from before 1960 seems generally higher than the flow since 1960, suggesting that the flow has decreased over time. Reasons for the observed change in summary statistics for the flow values could range from a variety of environmental or anthropocentric causes that would be interesting to investigate. Furthermore, for question 2, it was interesting to know that the minimum and maximum flow types do correspond to seasons. The maximum values correspond to spring and summer seasons, while the minimum values correspond to spring, autumn, and winter seasons. Since the station location that collected this data is unknown to me, knowing that the flow changes with the seasons helps me better understand potential locations of the station, and the subsequent environment that influence the flow in that region. 


# Task 2: Tidy your data

### 2.1 Is flow_sample tidy?

Flow_sample data set is tidy because each column is a variable: station_id, year, month, day, flow, and extreme_type. Within each row of the data set, there is an observation from the water gauge that shows when the data was collected and what the flow rate was at that observation. Furthermore, each cell is a single value that identifies the observation for each variable. Each cell only contains one value, and that value is reflective of both the information from the observation, and the variable from the column header. The cell does not contain more than 1 value. Since the data set matches all 3 criteria, I believe the flow_sample data set is tidy. 

### 2.2 Untidy and Tidy Data

Since the flow_sample data set is tidy, I will untidy it. 

Untidy the table to categorize each flow measurement by the extreme_type that the data was categorized as. The reason I have untidy-ed the data set in this manner is to have separate the maximum and minimum flow values, so we can more easily distinguish see which flow values are considered in the individual extreme types.

```{r}
# Data set before untidy
head(flow_sample)

# Expand the extreme types into separate columns
untidy <- flow_sample %>% 
  pivot_wider(names_from = extreme_type, values_from = flow)

# Data set after untidy
head(untidy)
```

Now, I will re-tidy the data set to make the minimum and maximum variable entries into one variable named "extreme_type"
```{r}
# Merge the Maximum and Minimum columns into one variable named Extreme Type
tidy <- untidy %>% 
  pivot_longer(cols = c(-station_id, -year, -month, -day, -sym), 
               names_to = "extreme_type", 
               values_to = "flow")

# Data set after re-tidy
head(tidy)
```

### 2.3 Choose questions for Milestone 3

Questions for Milestone 3 are:

1. How does the pattern of flow change over the seasons?
2. Are there changes in flow patterns by decade?

Question 1 is modified from Milestone 2 (this milestone's) Question #3, as I plan to broaden my research to also include investigation of minimum flow patterns over the seasons. To gain a more comprehensive view of the data and how the flow changes over time or seasonal patterns, I should consider both maximum and minimum flows. Therefore, I have chosen to continue with this question to work on the patterns seen in minimum flow values.

Question 2 is directly from the 4th question in Milestone 2, but without the specification of maximum flow values. This question is broad enough to continue working on finding patterns within the data and calculating more summary statistics within and between each decade of flow, and for each extreme type. 

The version of data I will be using will include new columns for decade and season. Since I have already performed the operations for maximum flow values, I will now filter for minimum values. 

```{r}
flow_sample_MDA3 <- flow_sample %>% 
   select(station_id, year, extreme_type, month, day, flow) %>%
   filter(extreme_type == 'minimum') %>%
   mutate(season = case_when(month < 3 ~ "winter",
                                 month < 6 ~ "spring",
                                 month < 9 ~ "summer",
                                 month < 12 ~ "autumn",
                                 TRUE ~ "winter")) %>%
   mutate(decade = case_when(year > 2009 ~ "2010_2019",
                            year > 1999 ~ "2000_2009",
                            year > 1989 ~ "1990_1999",
                            year > 1979 ~ "1980_1989",
                            year > 1969 ~ "1970_1979",
                            year > 1959 ~ "1960_1969",
                            year > 1949 ~ "1950_1959",
                            year > 1939 ~ "1940_1949",
                            year > 1929 ~ "1930_1939",
                            year > 1919 ~ "1920_1929",
                            year > 1909 ~ "1910_1919",
                            year > 1899 ~ "1900_1909"))
print(flow_sample_MDA3)

```
